# Analysis of longitudinal data
Hoo boy. It's the last week of IODS course, and this time we're analyzing data
that includes non-independent measurements (e.g. repeated measurements of
a test subject over time). Let's start with the libraries and data exploration!
```{r ch6-setup}
```

This week's assignment has a twist! We are going to do the exercises from 
Vehkalahti and Everitt (2019) book *Multivariate Analysis for The Behavioral Sciences*, 
chapters 8 and 9, except with the data sets swapped between the chapters. So first
we have the `rats` data introduced in Ch. 9: ***[[TODO]]***

> To begin to see how linear mixed effects models are applied in practice, we
shall use some data from a nutrition study conducted in three groups of rats
(Crowder and Hand, 1990). The three groups were put on different diets, and
each animal’s body weight (grams) was recorded repeatedly (approximately
weekly, except in week seven when two recordings were taken) over a 9-week
period. The question of most interest is whether the growth profiles of the
three groups differ. The data are shown in Table 9.1.

```{r c6-data}
```

Next, we have the `bprs`data introduced in Ch. 8: ***[[TODO]]***

> Table 8.1 taken from Davis (2002). Here 40 male subjects were
randomly assigned to one of two treatment groups and each subject was rated
on the brief psychiatric rating scale (BPRS) measured before treatment be-
gan (week 0) and then at weekly intervals for eight weeks. The BPRS assesses
the level of 18 symptom constructs such as hostility, suspiciousness, halluci-
nations and grandiosity; each of these is rated from one (not present) to seven
(extremely severe). The scale is used to evaluate patients suspected of having
schizophrenia.
