# Analysis of longitudinal data
Hoo boy. It's the last week of IODS course, and this time we're analyzing data
that includes non-independent measurements (e.g. repeated measurements of
a test subject over time). Let's start with the libraries and data exploration!
```{r ch6-setup, message=FALSE}
# Setup block
# Remove the '##' from each line of R output blocks
knitr::opts_chunk$set(comment = '')

# Libraries
library(dplyr)
library(ggplot2)
library(stringr)
library(lme4)
```

This week's assignment has a twist! We are going to repeat the examples from 
Vehkalahti and Everitt (2019) book *Multivariate Analysis for The Behavioral Sciences*, 
chapters 8 and 9, except with the data sets swapped between the chapters. I've 
wrangled the data to a "long table" format in the corresponding wrangling exercise.
You can find the code [here](https://github.com/pkars/IODS-project/blob/master/data/meet_and_repeat.R).

So first we have the `rats` data introduced in Ch. 9. The authors describe the data as following:

*[W]e shall use some data from a nutrition study conducted in three groups of rats (Crowder and Hand, 1990). The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ.*

```{r c6-data-rats}
rats <- read.csv("data/rats.csv")
rats$ID <- as.factor(rats$ID)
rats$Group <- as.factor(rats$Group)
str(rats)
summary(rats)
```
There are 16 rats in total, 8 in group 1, 4 in group 2 and 4 in group 3. Each rat has an
unique ID. `time` column is the measurement day.

Next, we have the `bprs`data introduced in Ch. 8. The authors' description follows:

*Table 8.1 taken from Davis (2002). Here 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.*

```{r c6-data-bprs}
bprs <- read.csv("data/bprs.csv")
# Create ID column before factor conversion
# treatment "1" and subject "1" -> 101
# treatment "2" and subject "20" -> 220
bprs$ID <- as.factor(bprs$treatment * 100 + bprs$subject)
# Categorical integers to categories...
bprs$treatment <- as.factor(bprs$treatment)
bprs$subject <- as.factor(bprs$subject)
str(bprs)
summary(bprs)

# Check that there are 40 unique ID values
unique(bprs$ID)
```
So, that's 40 subjects divided evenly into two groups. This time however, the subject
IDs are not unique within the data. They are unique within the group, though. That
will cause me some issues later on, ~~because I suck at R~~ so I've created unique
ID column here. In hindsight, this should've been done at the data wrangling stage, 
though I don't recall it being in the instructions.

## Working with RATS data

### Plot of the rats' weights (Fig. 8.1).
So here is a graphical overview of the time series of rats' weight development. 
Group 1 seems to consist of smaller rats than the other two groups. 
Also there is one huge rat in group 2, which will skew the mean value for that group.
```{r}
ggplot(rats, aes(x = time, y = weight, linetype= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, 2)) +
  facet_grid(. ~ Group, labeller=label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(rats$weight), max(rats$weight)))

```

### Standardized weights (Fig 8.2)
Next up is "tracking", i.e. plotting the time series of standardized weights. That'll
show how a rat gains weight compared to the others. This feels like anomaly detection
in atmospheric dynamics, or de-trending. Though those things are a bit different
-- or at least the questions they try to answer are different. I think. Anyways,
now the lines are not always increasing, so the relative development is easier
to see.
```{r}
rats_tracked <- rats %>% 
  group_by(time) %>%
  mutate(std.weight = (weight - mean(weight))/sd(weight)) %>%
  ungroup()

ggplot(rats_tracked, aes(x = time, y = std.weight, linetype= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, 2)) +
  facet_grid(. ~ Group, labeller=label_both) +
  theme(legend.position = "none") +
  scale_y_continuous("Standardized weight")

```

### Mean response profiles (Fig. 8.3)
```{r}
ratss <- rats %>%
  group_by(Group, time) %>%
  summarise( mean = mean(weight), se = sd(weight)/sqrt(n()), .groups="keep") %>%
  ungroup()

# Glimpse the data
glimpse(ratss)

# Plot the mean profiles
ggplot(ratss, aes(x = time, y = mean, linetype = Group, shape="1", color=Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3, show.legend = F) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

### Boxplots for rats data (Fig. 8.4)
```{r}
ggplot(rats, aes(x=as.factor(time), y=weight, color=Group)) +
  geom_boxplot()
```

### Boxplots of mean summary measures for the three rat groups (Fig. 8.5)
```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline day 1).
rats_sm <- rats %>%
  filter(time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(weight), .groups="keep" ) %>%
  ungroup()

# Glimpse the data
glimpse(rats_sm)

# Draw a boxplot of the mean versus treatment
ggplot(rats_sm, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), time > 1")
```

### Same as above but with outliers (if any) removed (Fig. 8.6)
Okay so, group 2 has an outlier. Remove it. Though I'm removing 1 rat from a total 4 rats,
meaning that the statistics start to be *quite* dubious... `n=3` doesn't sound like
a large enough of a sample.
```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
rats_sm1 <-  rats_sm %>% filter(mean < 575)
glimpse(rats_sm1)

ggplot(rats_sm1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight), time > 1")
```

### *t*-test for Mean Summary Measure (table 8.3)
*t*-test is a paired test, so there can be only two variables. Luckily three 
groups equate to three pairs, so I can do the t-test for each pair with minimal effort (copypaste).
```{r}
rats_sm1_noG3 <- rats_sm1 %>% filter(Group != 3)
t.test(mean ~ Group, data = rats_sm1_noG3, var.equal = TRUE)

rats_sm1_noG2 <- rats_sm1 %>% filter(Group != 2)
t.test(mean ~ Group, data = rats_sm1_noG2, var.equal = TRUE)

rats_sm1_noG1 <- rats_sm1 %>% filter(Group != 1)
t.test(mean ~ Group, data = rats_sm1_noG1, var.equal = TRUE)
```

### Analysis of Covariance (ANOVA) (table 8.4)
For the ANOVA I need a baseline for each row -- and that is easy to pick up from
the original data. This works because the `rats_sm` contains only one row for each
rat -- it is a summary of all measurements -- and the original data has one row
for each rat, too. If it weren't for the summarized data, I'd have to join by
ID columns. Although that would be clearer than this, but (probably) slower...
```{r}
original_rats <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt",
                          header=TRUE, sep="\t")
rats_sm2 <- rats_sm %>%
  mutate(baseline = original_rats$WD1)

fit <- lm(mean ~ baseline + Group, data = rats_sm2)
anova(fit)
```

Lastly, I'm omitting the Table 8.6. It doesn't use the BPRS data in the original,
so strictly speaking cannot swap that data to RATS dataset. It also demonstrates
how to deal with missing data, so performing the analysis with a full dataset
is not the most productive way (of course I could randomly remove some of the 
data and work that way, but I'm being lazy here).

## Working with BPRS data

Moving on to the BPRS data and Chapter 9 examples. I might take some liberties
here and there in visualising the data instead of 100% faithfully replicating
the figures. Mainly because the data is different so I might need to adjust 
the visuals a bit to increase readability.

### Week vs BPRS with group numbers (Fig. 9.1)
Here the two groups overlap more than the rats, and we have more individuals here as well!

```{r}
ggplot(bprs, aes(x=week, y=bprs, color=treatment)) +
  geom_text(aes(label=treatment), show.legend=FALSE)
```

### Results from fitting linear reg.model ignoring the repeated-measures structure of the data (Table 9.3)
```{r}
# create a regression model RATS_reg
bprs_reg <- lm(bprs ~ week + treatment, data = bprs)
# print out a summary of the model
summary(bprs_reg)
```

I'm going to cheat a bit here and plot an extra figure or two. Because I want to
see how this faulty analysis looks like!

```{r}
fit <- fitted(bprs_reg, bprs)

ggplot(bprs, aes(x=week, y=bprs)) +
  geom_line(aes(group=subject), color="gray75") +
  geom_point() +
  geom_smooth(method="lm") +
  labs(title="BPRS by week (with regression) for both treatment groups") +
  facet_wrap(vars(treatment))

```

Since the `treatment` is a categorical value, I can plot the linear regression
this way. It helps that there are two categories in `treatment`, though. I included
the the time series as lines, too, to highlight why this sort of regression modeling
doesn't work here. Okay, sure, it shows the downward trend in both groups, but 
for analysing trends you'd want to consider the "repeated measures" nature of the
data anyways[^4].

### Individual BTRS profiles (Fig. 9.2)
Here's where the custom `ID` column comes in handy. I need to be able to differentiate
between the subjects with the same `subject` value but different `treatment` value.
Otherwise this plot would become confused, poor thing.
```{r}
# Check that the ID column is included
str(bprs)

# Plot the RATSL data
ggplot(bprs, aes(x = week, y = bprs, group=ID)) +
  geom_line(aes(linetype=treatment)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```

### Scatterplot matrix of repeated measures (Fig. 9.3)
This thing requires the original data. Maybe I could pivot the data... Yep, that works.
```{r}
bprs_w <- tidyr::pivot_wider(bprs, names_from=weeks, values_from=bprs, 
                             id_cols=c("treatment", "subject", "ID")) %>% 
  as.data.frame
str(bprs_w)

plot(bprs_w[-1:-3])
```

### Results from fitting random intercept model (Table 9.4)
```{r}
# Create a random intercept model
bprs_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = bprs, REML = FALSE)

# Print the summary of the model
summary(bprs_ref)

```

### Results from fitting random intercept and slope model (Table 9.5)
```{r}
# create a random intercept and random slope model
bprs_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = bprs, REML = FALSE)

# print a summary of the model
summary(bprs_ref1)

# perform an ANOVA test on the two models
anova(bprs_ref1, bprs_ref)

```

So, as there are duplicate values in `subject` column, I'm not so sure if that
fitting did what we wanted to do here. Let's do this again using `ID` instead of `subject`.
```{r}
# Create a random intercept model
bprs_ref_id <- lmer(bprs ~ week + treatment + (1 | ID), data = bprs, REML = FALSE)

# Print the summary of the model
summary(bprs_ref)

# create a random intercept and random slope model
bprs_ref_id1 <- lmer(bprs ~ week + treatment + (week | ID), data = bprs, REML = FALSE)

# print a summary of the model
summary(bprs_ref_id1)

# perform an ANOVA test on the two models
anova(bprs_ref_id1, bprs_ref_id)
```

### Results from Fitting the Random Intercept and Slope Model that Allows for a Group×Time Interaction to Rat Growth Data (Table 9.6)
```{r}
# create a random intercept and random slope model with the interaction
bprs_ref2 <- lmer(bprs ~ week * treatment + (week | ID), data = bprs, REML = FALSE)

# print a summary of the model
summary(bprs_ref2)

# perform an ANOVA test on the two models
anova(bprs_ref2, bprs_ref1)


```

### Fitted and observed growth profiles (Figure 9.4)
```{r}
# draw the plot of RATSL with the observed Weight values
ggplot(bprs, aes(x = week, y = bprs, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  #scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  #scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(bprs_ref2, bprs)

# Create a new column fitted to RATSL
bprs2 <- bprs
bprs2$Fitted <- Fitted

# draw the plot of RATSL with the Fitted values of weight
ggplot(bprs2, aes(x = week, y = Fitted, group = ID)) +
  geom_line(aes(linetype = treatment)) +
  #scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  #scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")
```


[^4]: I'm having trouble imagining what sort of measures would be time series
measures *and simultaneously not repeated* measures, as "time series" already
implies repetition. Perhaps sequences of truly random noise, such as atmospheric noise,
could be considered as such measure?
