---
title: "IODS course project"
author: "Petteri Karsisto"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Course Project

<!-- This will include Chapter 1 (that is updated in its own file) in the document.-->
```{r 'ch1', child = "chapter1.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>
  
***

<!-- Chapter 2 -->  
```{r 'ch2', child = "chapter2.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>

***

<!-- Chapter 3 -->
```{r 'ch3', child = "chapter3.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>

***

<!-- Chapter 4 -->
```{r 'ch4', child = "chapter4.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>

***

<!-- Chapter 5 -->
```{r 'ch5', child = "chapter5.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>

***

<!-- Chapter 6 -->
```{r 'ch6', child = "chapter6.Rmd"}
```

<a href="#top" class="return-link">Back to top.</a>

***

# Final word

So, this was it! That was the last chapter of this course and with it this course
diary comes to an end. In a proper [book-ends](https://tvtropes.org/pmwiki/pmwiki.php/Main/Bookends)-style,
I'm going to write a bit about my thoughts about the course.

First of all, I had fun! While the course had a lot of work and strict deadlines,
I still enjoyed writing this diary. Partially because I decided to write in my
own style -- with the casual remarks and all. I wanted to make the writing fun for me
and reading enjoyable to the readers. Based on the peer reviews I got along the way,
I'd say that I was successful!

Second, the R language was a new experience for me. The tasks weren't too difficult
especially when the Datacamp exercises pretty much solve nearly everything on the
coding side. The hardest part was overcoming my initial way of solving things --
R is not Python, and things work differently here! Once I realised this the coding
became smoother.

Third, I learned a lot about data analysis! I hope I can use this new knowledge
as a basis when I start delving deeper into my research (which will inevitably require
a better understanding of statistics).

Finally, thanks for Kimmo Vehkalahti, Reijo Sund, and the teaching assistants for
organising the course and you, dear reader, for reading my ramblings.

Petteri signing out.

<a href="#top" class="return-link">Back to top.</a>